---
title: "survival analysis"
author: "Wendy Zhang"
date: "2023-11-28"
output: html_document
---
Read in all three files:

```{r}
RNA_data <- read.csv("RNAseq_LUSC.csv")
```


```{r}
data_clinical_patient <- read.csv("data_clinical_patient.txt", sep = '\t', header = TRUE, skip = 4) 
#ski first 4 data that not relates to patient ID

data_mutations <- read.csv("data_mutations.txt", sep = '\t', header = TRUE )

```

Based on the Oncoplot matrix, do survival analysis on two cluster

```{r}
data_clinic_common<-subset(data_clinical_patient,PATIENT_ID %in% unique_pat)
data_clinic_filtered <- data_clinic_common[data_clinic_common$SEX != "", ]
data_clinic_filtered <- data_clinic_common[data_clinic_common$AJCC_PATHOLOGIC_TUMOR_STAGE != "", ]
```

Find the clustered patient:

```{r}
# Extract patient IDs for Cluster 1
cluster_1_patients <- row.names(cluster)[cluster$`cutree(res$tree_col, k = 2)` == 1]

#Get first 12 char
cluster_1_pat<-substr(cluster_1_patients, start = 1, stop = 12)

# Filter clinical data based on Cluster 1 patients
data_clinic_filtered_1<- data_clinic_filtered[data_clinic_filtered$PATIENT_ID %in% cluster_1_pat, ]
```

```{r}
# Extract patient IDs for Cluster 2
cluster_2_patients <- row.names(cluster)[cluster$`cutree(res$tree_col, k = 2)` == 2]

#Get first 12 char
cluster_2_pat<-substr(cluster_2_patients, start = 1, stop = 12)

# Filter clinical data based on Cluster 2 patients
data_clinic_filtered_2<- data_clinic_filtered[data_clinic_filtered$PATIENT_ID %in% cluster_2_pat, ]
```

Library needed:
```{r}
library("TCGAbiolinks")
library("survival")
library("survminer")
```

```{r}
library("SummarizedExperiment")
```
**CLUSTER 1**
```{r}
colnames(data_clinic_filtered_1)
```
```{r}
colnames(RNA_data)
```



Overall survival status
```{r}
# Create a boolean variable for deceased
data_clinic_filtered_1$deceased <- data_clinic_filtered_1$OS_STATUS == "1:DECEASED"

# Create an overall survival variable
data_clinic_filtered_1$overall_survival <- ifelse(data_clinic_filtered_1$deceased,
                                                  data_clinic_filtered_1$OS_MONTHS,
                                                  data_clinic_filtered_1$DAYS_LAST_FOLLOWUP)

# Create a survival object
surv_os <- Surv(time = data_clinic_filtered_1$overall_survival, 
                event = data_clinic_filtered_1$deceased)

# Fit the Kaplan-Meier estimator for OS
km_fit_os <- survfit(surv_os ~ 1)

plot(km_fit_os, main = "Kaplan-Meier Survival Curve", xlab = "Time (Months)", ylab = "Survival Probability - OS", col = 1)



```


How does gender affect survival?

```{r}
fit1<- survfit(surv_os ~ SEX, data = data_clinic_filtered_1)

# Print the fitted survival model
print(fit1)

# Plot the Kaplan-Meier survival curve with risk table
library(survminer)
ggsurvplot(fit1, data = data_clinic_filtered_1, pval = TRUE, risk.table = TRUE, risk.table.col = "strata", risk.table.height = 0.35)
```

The p-value is non-significant, so gender alone does not significantly sway prognosis.


How does tumor stage affect survival?
```{r}
# Remove "A", "B", or "C" at the end of the tumor stage name
data_clinic_filtered_1$tumor_stage <- gsub("[ABC]$", "", data_clinic_filtered_1$AJCC_PATHOLOGIC_TUMOR_STAGE)

# Remove those with stage "not reported"
data_clinic_filtered_1$tumor_stage[data_clinic_filtered_1$AJCC_PATHOLOGIC_TUMOR_STAGE == ""] <- NA

# Check the updated distribution of tumor stages
table(data_clinic_filtered_1$tumor_stage)

```

```{r}
# Fit the Kaplan-Meier estimator by tumor stage
fit2 = survfit(Surv(overall_survival, deceased) ~ tumor_stage, data=data_clinic_filtered_1)

# Extract the survival p-value
pval2 <- surv_pvalue(fit2, data = data_clinic_filtered_1)$pval
print(pval2)

# Plot the Kaplan-Meier survival curve with risk table
library(survminer)
ggsurvplot(fit2, data = data_clinic_filtered_1, pval = TRUE, risk.table = TRUE, risk.table.height = 0.35)
```
The p-value is significant, so different tumor stages does significantly sway prognosis.

If we set significance level alpha to 0.05, then this p-value of 0.039 is less than 0.05, so we are rejecting the null hypothesis.
The rejection of the null hypothesis indicates that there is evidence in the data to suggest that the survival curves for different tumor stages are different. Therefore, statistically there is association between tumor stage and survival.




**CLUSTER 2**

```{r}
# Create a boolean variable for deceased
data_clinic_filtered_2$deceased <- data_clinic_filtered_2$OS_STATUS == "1:DECEASED"

# Create an overall survival variable
data_clinic_filtered_2$overall_survival <- ifelse(data_clinic_filtered_2$deceased,
                                                  data_clinic_filtered_2$OS_MONTHS,
                                                  data_clinic_filtered_2$DAYS_LAST_FOLLOWUP)

# Create a survival object
surv_os_2 <- Surv(time = data_clinic_filtered_2$overall_survival, 
                event = data_clinic_filtered_2$deceased)

# Fit the Kaplan-Meier estimator for OS
km_fit_os_2 <- survfit(surv_os_2 ~ 1)

plot(km_fit_os_2, main = "Kaplan-Meier Survival Curve", xlab = "Time (Months)", ylab = "Survival Probability - OS", col = 1)

```

How gender affect?

```{r}
fit3 <- survfit(surv_os_2 ~ SEX, data = data_clinic_filtered_2)

# Print the fitted survival model
print(fit3)

# Plot the Kaplan-Meier survival curve with risk table
library(survminer)
ggsurvplot(fit3, data = data_clinic_filtered_2, pval = TRUE, risk.table = TRUE, risk.table.col = "strata", risk.table.height = 0.35)
```
The p-value is non-significant, so gender alone does not significantly sway prognosis.


How does tumor stage affect survival?
```{r}
# Remove "A", "B", or "C" at the end of the tumor stage name
data_clinic_filtered_2$tumor_stage <- gsub("[ABC]$", "", data_clinic_filtered_2$AJCC_PATHOLOGIC_TUMOR_STAGE)

# Remove those with stage "not reported"
data_clinic_filtered_2$tumor_stage[data_clinic_filtered_2$AJCC_PATHOLOGIC_TUMOR_STAGE == ""] <- NA

# Check the updated distribution of tumor stages
table(data_clinic_filtered_2$tumor_stage)

```


```{r}
# Fit the Kaplan-Meier estimator by tumor stage
fit4 = survfit(Surv(overall_survival, deceased) ~ tumor_stage, data=data_clinic_filtered_2)

# Extract the survival p-value
pval4 <- surv_pvalue(fit4, data = data_clinic_filtered_2)$pval
print(pval4)

# Plot the Kaplan-Meier survival curve with risk table
library(survminer)
ggsurvplot(fit4, data = data_clinic_filtered_2, pval = TRUE, risk.table = TRUE, risk.table.height = 0.35)
```

